---
title: "Data Visualization"
author: "Wei-Hung Weng"
date: "Nov 3, 2016"
output: ioslides_presentation
subtitle: "BMI701 Lab Session 7"
highlighter: highlight.js
---

# Plotting System

## Plotting Systems in R - Base
- "Artist's palette" model
- Start with blank canvas and build up from there
- Start with `plot` function (or similar)
- Use annotation functions to add/modify (`text`, `lines`, `points`, `axis`)

## Plotting Systems in R - Base
- Convenient, mirrors how we think of building plots and analyzing data
- Can't go back once plot has started (i.e. to adjust margins); need to plan in advance
- Difficult to 'translate' to others once a new plot has been created (no graphical 'language')
- Plot is just a series of R commands

## Plotting Systems in R - Lattice
- Plots are created with a single function call (`xyplot`, `bwplot`, etc.)
- Most useful for conditioning types of plots: Looking at how $y$ changes with $x$ across levels of $z$
- Things like margins/spacing set automatically because entire plot is specified at once
- Good for putting many many plots on a screen

## Plotting Systems in R - Lattice
- Sometimes awkward to specify an entire plot in a single function call
- Annotation in plot is not intuitive
- Use of panel functions and subscripts difficult to wield and requires intense preparation
- Cannot 'add' to the plot once it's created

## Plotting Systems in R - ggplot2
- Split the difference between `base` and `lattice`
- Automatically deals with spacings, text, titles but also allows you to annotate by adding `+`
- Superficial similarity to lattice but generally easier/more intuitive to use
- Default mode makes many choices for you (but you _can_ customize!)

# ggplot2

## What is ggplot2?
- An implementation of `The Grammar of Graphics` by Leland Wilkinson
- Written by `Hadley Wickham` (while he was a graduate student at Iowa State)
- A 'third' graphics system for R (along with `base` and `lattice`)
- [Web site](http://ggplot2.org)

## What is ggplot2?
- `Grammar of graphics` represents an abstraction of graphics ideas/objects
- Think 'verb', 'noun', 'adjective' for graphics
- Allows for a 'theory' of graphics on which to build new graphics and graphics objects

## Grammer of Graphics (from `ggplot2` book)
### "In brief, the grammar tells us that a statistical graphic is a `mapping` from data to `aesthetic` attributes (colour, shape, size) of `geometric` objects (points, lines, bars). The plot may also contain statistical transformations of the data and is drawn on a specific coordinate system"

## Basic Components of a ggplot2 Plot (Grammar)
- `data frame`
- `aesthetic mappings`: how data are mapped to color, size, thickness, ...
- `geoms`: geometric objects like points, lines, shapes
- `facets`: for conditional plots (subgraph)
- `stats`: statistical transformations like binning, quantiles, smoothing. 
- `scales`: what scale an aesthetic map uses (example: male = red, female = blue)
- `coordinate system` 

## The Basics: `qplot()`
- Works much like the `plot` function in base graphics system
- Looks for data in a data frame, similar to lattice
- Plots are made up of `aesthetics` (size, shape, color) and `geoms` (points, lines)

## The Basics: `qplot()`
- Factors are important for indicating subsets of the data; they should be `labeled`
- The `qplot()` hides what goes on underneath, which is okay for most operations
- `ggplot()` is the core function and very flexible for doing things `qplot()` cannot do

## ggplot2 "Hello, world!"
```{r, fig.height=4, warning=F, message=F}
library(ggplot2)
# qplot(x, y, data) -> scatter plot
qplot(carat, price, data=diamonds)
```

## Modifying aesthetics
```{r, fig.height=4}
qplot(carat, price, data=diamonds, color=clarity)
```

## Adding a geom
```{r, fig.height=4}
# add smooth curve
qplot(carat, price, data=diamonds, 
      geom=c("point", "smooth")) 
```

## Histograms
```{r, fig.height=3.5}
qplot(price, data=diamonds, fill=cut)
```

## Facets
```{r, fig.height=3.5}
# facets = . ~ cut (horizontal)
qplot(carat, price, data=diamonds, 
      facets=.~cut)
```

## Facets
```{r, fig.height=4}
# facets = cut ~ . (vertical)
qplot(carat, price, data=diamonds, 
      facets=cut~.)
qplot(hwy, data = mpg, facets = drv ~ ., binwidth = 2)
```

## Summary of qplot()
- The `qplot()` function is the analog to `plot()` but with many built-in features
- Syntax somewhere in between `base/lattice`
- Produces very nice graphics, essentially publication ready (if you like the design)
- Difficult to go against the grain/customize (don't bother; use full ggplot2 power in that case)

## Building Plots with ggplot2
- When building plots in ggplot2 (rather than using qplot) the "artist’s palette" model may be the closest analogy
- Plots are built up in layers
  - Plot the data
  - Overlay a summary
  - Metadata and annotation

## Annotation
- Labels: `xlab()`, `ylab()`, `labs()`, `ggtitle()`
- Each of the `geom` functions has options to modify 
- For things that only make sense globally, use `theme()` 
  - Example: `theme(legend.position = "none")` 
- Two standard appearance themes are included
  - `theme_gray()`: The default theme (gray background)
  - `theme_bw()`: More stark/plain 

## Other themes 
- `ggthemes` packages [Website](https://cran.r-project.org/web/packages/ggthemes/vignettes/ggthemes.html)
- `xkcd` packages [Website](http://xkcd.r-forge.r-project.org/)

## Scatter plot in ggplot2 geom_point()
```{r, fig.height=4.5, fig.height=3}
# aes: Aesthetic attributes, color, shape, size, thickness
# geom_*: Geometric objects, point, line, boxplot, bar
ggplot(diamonds, aes(x = carat, y = price)) + 
  geom_point()
```

## Box plot in ggplot2 geom_boxplot()
```{r, fig.height=4.5, fig.height=3}
# aes: Aesthetic attributes, color, shape, size, thickness
# geom_*: Geometric objects, point, line, boxplot, bar
ggplot(diamonds, aes(x = cut, y = price)) + 
  geom_boxplot()
```

## Faceting in ggplot2 facet_grid()
```{r, fig.height=4.5, fig.height=3}
# facet_grid: add subgraph
ggplot(diamonds, aes(x = carat, y = price)) + 
  geom_point() +
  facet_grid(cut~.)
```


## Faceting in ggplot2 facet_grid()
```{r, fig.height=4.5, fig.height=3}
# facet_grid: add subgraph
ggplot(diamonds, aes(x = carat, y = price)) + 
  geom_point() +
  facet_grid(.~cut)
```

## Regression line in ggplot2 geom_smooth()
```{r, fig.height=4.5, fig.height=3}
# geom_smooth: method='lm' (linear regression)
ggplot(diamonds, aes(x = carat, y = price)) + 
  geom_point() +
  facet_grid(cut~.) + 
  geom_smooth(method='lm')
```

## Color group in ggplot2 color=？
```{r, fig.height=4.5, fig.height=3}
# color = cut
ggplot(diamonds, aes(x = carat, y = price, color = cut)) + 
  geom_point() +
  geom_smooth(method='lm')
```

## A Note about Axis Limits
```{r, fig.height=4.5, fig.height=3}
testdat <- data.frame(x = 1:100, y = rnorm(100))
testdat[50, 2] <- 100  # Outlier!
plot(testdat$x, testdat$y, type = "l", ylim = c(-3,3))
```

## A Note about Axis Limits
```{r, fig.height=4.5, fig.height=3}
g <- ggplot(testdat, aes(x = x, y = y))
g + geom_line()
```

## Axis Limits
```{r, fig.height=4.5}
g + geom_line() + ylim(-3, 3)
```

## Axis Limits
```{r, fig.height=4.5}
g + geom_line() + coord_cartesian(ylim = c(-3, 3))
```

# Application

## Error bar
- Standard deviation (SD)
- Standard error (SE)
- Confidence interval (CI)
    
## Bar Chart without Error Bar in ggplot2
```{r, fig.height=3.5}
library(datasets)
airquality$Month <- as.factor(airquality$Month)
airquality.mean <- aggregate(Ozone~Month, airquality, mean)
ggplot() + 
  geom_bar(data=airquality.mean, aes(x=Month, y=Ozone), stat="identity")
```

## Bar Chart with Error Bar in ggplot2
```{r, fig.height=3}
airquality.sd <- aggregate(Ozone~Month, airquality, sd)
airquality.eb <- merge(airquality.mean, airquality.sd, by="Month")
ggplot(data=airquality.eb) +
  geom_bar(aes(x=Month, y=Ozone.x), stat = "identity")+
  geom_errorbar( # Ozone.x = mean, Ozone.y = sd
    aes(x=Month, ymin=Ozone.x-Ozone.y, ymax=Ozone.x+Ozone.y), width=.1)
```

## `ggplot2` Reference
- The `ggplot2` book by Hadley Wickham
- The `R Graphics Cookbook` by Winston Chang (examples in base plots and in ggplot2)
- [ggplot2 web site](http://ggplot2.org)
- [ggplot2 doc](http://docs.ggplot2.org/current/)
- [ggplot cheat sheet](https://www.rstudio.com/wp-content/uploads/2015/03/ggplot2-cheatsheet.pdf)
- [r blooger](http://www.r-bloggers.com/basic-introduction-to-ggplot2/)

# Map

## Heatmap
[Reference](http://flowingdata.com/2010/01/21/how-to-make-a-heatmap-a-quick-and-easy-solution/)
```{r message=FALSE, warning=F}
nba <- read.csv("http://datasets.flowingdata.com/ppg2008.csv")
head(nba)
```

## Heatmap (Wide to Long)
```{r message=FALSE, warning=F}
library(reshape2) # for melt()
nba.m <- melt(nba, id.vars = "Name") # wide to long based on name
head(nba.m, 10)
```

## Heatmap geom_tile()
```{r message=FALSE, warning=F, fig.height=4}
library(ggplot2) # for ggplot()
ggplot(nba.m, aes(variable, Name)) +
    geom_tile(aes(fill = value), colour = "white") +
    scale_fill_gradient(low = "white", high = "steelblue")
```

## Heatmap: scale
```{r message=FALSE, warning=F, fig.height=4}
head(nba, 2)
nba[, 2:21] <- apply(nba[, 2:21], 2, scale) # scale, average to 0
head(nba, 2)
```

## Heatmap: scale
```{r message=FALSE, warning=F, fig.height=4}
nba.m <- melt(nba)
ggplot(nba.m, aes(variable, Name)) +
    geom_tile(aes(fill = value),colour = "white") +
    scale_fill_gradient(low = "white",high = "steelblue")
```

## Treemap
- Use `treemapify` packages
- [Reference](https://github.com/wilkox/treemapify)
- [G20 Dataset](https://en.wikipedia.org/wiki/G-20_major_economies)
```{r echo=F, message=FALSE, warning=F, fig.height=4}
if (!require('treemapify')){
    library(devtools)
    install_github("wilkox/treemapify")
    library(treemapify)
}
data(G20)
treeMapCoordinates <- treemapify(data=G20,
    area = "Nom.GDP.mil.USD",fill = "HDI",
    label = "Country",group = "Region")
ggplotify(treeMapCoordinates)+ 
    scale_fill_gradient(low = "white",high = "steelblue")
```

## Treemap - Data Processing 
```{r message=FALSE,warning=F,fig.height=4}
if (!require('treemapify')){
    library(devtools)
    install_github("wilkox/treemapify")
    library(treemapify)
}
data(G20)
head(G20)
```

## Treemap - Parameter
```{r message=FALSE, warning=F, fig.height=4}
# treemapify: convert format
# area, fill, label, group
treeMapCoordinates <- treemapify(data=G20,
    area = "Nom.GDP.mil.USD", fill = "HDI",
    label = "Country", group = "Region")
head(treeMapCoordinates)
```

## Treemap - Plotting
```{r message=FALSE, warning=F, fig.height=4}
ggplotify(treeMapCoordinates)
```

## Treemap - Plotting
```{r message=FALSE, warning=F, fig.height=4}
ggplotify(treeMapCoordinates)+ 
    scale_fill_gradient(low = "white",high = "steelblue")
```

## Choropleth Map
- [Choropleth map?](https://en.wikipedia.org/wiki/Choropleth_map)
- Stat data to map
- `choroplethr` & `choroplethrMaps` packages
```{r, echo=F, warning=F,message=F}
.packages <- c("choroplethr", "choroplethrMaps")
.inst <- .packages %in% installed.packages()
if(length(.packages[!.inst]) > 0) suppressMessages(install.packages(.packages[!.inst], repos="http://cran.rstudio.com/"))
invisible(lapply(.packages, library, character.only=TRUE))

data(df_pop_state) # population
state_choropleth(df_pop_state) # population to map
```

## Taiwan
- `rgdal`, `rgeos`,`maptools` packages for map files
- `ggplot2` & `RColorBrewer` for plotting
```{r, echo=F, warning=F,message=F, fig.height=4}
.packages <- c("rgdal", "rgeos", "maptools", "ggplot2", "gpclib", "RColorBrewer")
.inst <- .packages %in% installed.packages()
if(length(.packages[!.inst]) > 0) suppressMessages(install.packages(.packages[!.inst], repos="http://cran.rstudio.com/"))
invisible(lapply(.packages, library, character.only=TRUE))
gpclibPermit()

tw_new <- readShapeSpatial("lab07_tw/Town_MOI_1041215.shp")

tw_new.df <- fortify(tw_new, region = "T_UID") # from ggplot2 package
mydata <- data.frame(NAME_2=tw_new$T_Name, id=tw_new$T_UID,
                     prevalence=rnorm(length(tw_new$T_UID)))
final.plot <- merge(tw_new.df, mydata,by="id",all.x=T)

twcmap<-ggplot() +
    geom_polygon(data = final.plot, 
                 aes(x = long, y = lat, group = group, 
                     fill = prevalence), 
                 color = "black", size = 0.25) + 
    coord_map()+ # fixed scale
    scale_fill_gradientn(colours = brewer.pal(9,"Reds"))+
    theme_void()+
    labs(title="Prevalence of X in Taiwan")
twcmap
```

## Taiwan
- Use [Taiwan open data](http://data.gov.tw/node/7441)
- Include longitude, latitute, boundary, ...
- `shapefile` [Wiki](https://zh.wikipedia.org/zh-tw/Shapefile)
- Spatial data format
- [Reference](http://www.r-bloggers.com/mapping-with-ggplot-create-a-nice-choropleth-map-in-r/)

## Read `shapefile`
- Use `maptools` package, `readShapeSpatial` function
```{r, warning=F,message=F, fig.height=3}
.packages <- c("rgdal", "rgeos", "maptools", "ggplot2", "gpclib", "RColorBrewer")
.inst <- .packages %in% installed.packages()
if(length(.packages[!.inst]) > 0) suppressMessages(install.packages(.packages[!.inst], repos="http://cran.rstudio.com/"))
invisible(lapply(.packages, library, character.only=TRUE))

tw_new <- readShapeSpatial("Taiwan/Town_MOI_1041215.shp")
names(tw_new)
```

## Process shapefile
- `rgdal`, `rgeos`,`maptools`
- `fortify`: `shapefile` $\rightarrow$ `data.frame`
- [Reference](http://www.r-bloggers.com/mapping-with-ggplot-create-a-nice-choropleth-map-in-r/)
```{r, warning=F,message=F, fig.height=3}
head(tw_new$Town_ID)
tw_new.df <- fortify(tw_new, region = "T_UID") # from ggplot2 package
```

## Process shapefile
- [Reference](http://www.r-bloggers.com/mapping-with-ggplot-create-a-nice-choropleth-map-in-r/)
```{r, warning=F,message=F, fig.height=3}
head(tw_new.df,10)
```

## Generate Data
```{r, warning=F,message=F, fig.height=3}
# Randomly generate some data for plotting
mydata <- data.frame(NAME_2=tw_new$T_Name, id=tw_new$T_UID,
                     prevalence=rnorm(length(tw_new$T_UID)))
head(mydata)
```

## Chinese Encoding
```{r, warning=F,message=F, fig.height=3}
# use iconv to convert
# from big5 to utf-8
mydata$NAME_2 <- iconv(as.character(mydata$NAME_2),
                       from="big5", to="UTF-8")
head(mydata,10)
```

## Merge Map and Data
```{r, warning=F, message=F, fig.height=3}
final.plot <- merge(tw_new.df, mydata,by="id", all.x=T)
head(final.plot,10)
```

## Plotting
```{r, warning=F, message=F, fig.height=3}
library(RColorBrewer)
twcmap <- ggplot() +
    geom_polygon(data = final.plot, 
                 aes(x = long, y = lat, group = group, 
                     fill = prevalence), 
                 color = "black", size = 0.25) + 
    coord_map()+
    scale_fill_gradientn(colours = brewer.pal(9,"Reds"))+
    theme_void()+
    labs(title="Prevalence of X in Taiwan")
```

## Plotting
```{r, warning=F,message=F, fig.height=5}
twcmap
```

## Google map
- `ggmap` package
```{r, echo=F, warning=F, message=F, fig.height=5}
if (!require('ggmap')){
    install.packages("ggmap")
    library(ggmap) # for get_map()
}
twmap <- get_map(location='Taiwan', zoom=7, language="en-US")
ggmap(twmap) 
```

## `ggmap` (import Google map)
```{r, warning=F, message=F, fig.height=3}
if (!require('ggmap')){
    install.packages("ggmap")
    library(ggmap)
}
twmap <- get_map(location='Taiwan', zoom=7, language="en-US")
# location: name or long/lat
# zoom: 2-20
# language: map language
```

## `ggmap` (plot)
```{r, warning=F, message=F, fig.height=4}
ggmap(twmap)
```

## `ggmap` (maptype)
```{r, warning=F, message=F, fig.height=3}
# maptype: "terrain", "terrain-background", "satellite",
# "roadmap","hybrid" (google maps), "terrain", "watercolor", 
# "toner" (stamen maps), 
# or a positive integer for cloudmade maps (see ?get_cloudmademap)
tpmap <- get_map(location = c(121.43, 24.93, 121.62, 25.19), 
                     zoom = 14, maptype = 'roadmap')
ggmap(tpmap, extent = 'device') # extent='device': full screen
```

## `ggmap` + choropleth
```{r, warning=F, message=F, fig.height=6}
ggmap(twmap) + # ggmap
    geom_polygon(data = final.plot,
        aes(x = long, y = lat, group = group, fill = prevalence), 
        color = "grey80", size = 0.1,alpha = 0.5) + 
  scale_fill_gradientn(colours = brewer.pal(9,"Reds"))
```

## Density Map - Using `ggplot2` + `ggmap`
```{r, echo=F, warning=F,message=F}
# Get location of each state
StateCenter <- data.frame( 
    region=tolower(state.name),
    lon=state.center$x,
    lat=state.center$y)
# Population data
StatePop <- merge(df_pop_state, StateCenter, by="region") 
# Value to point
PopPoint <- NULL 
for(i in 1:nrow(StatePop)){
    # 1M people -> one point
    for(j in 1:(StatePop[i, "value"] / 1000000)){
        PopPoint <- rbind(PopPoint, StatePop[i, ])   
    }
}

USMap <- get_map(location = "United States", zoom = 4)
densityMap <- ggmap(USMap, extent = "device") + 
  geom_density2d(data = PopPoint, aes(x = lon, y = lat), size = 0.3) +
  stat_density2d(data = PopPoint, 
                 aes(x = lon, y = lat, fill = ..level.., alpha = ..level..), 
                 size = 0.01, bins = 16, geom = "polygon") + 
  scale_fill_gradient(low = "green", high = "red", guide = FALSE) + 
  scale_alpha(range = c(0, 0.3), guide = FALSE)

densityMap
```

## Data Preprocessing
```{r warning=F, message=F}
# Get location of each state
StateCenter <- data.frame( 
    region=tolower(state.name),
    lon=state.center$x,
    lat=state.center$y)
head(StateCenter, 1)
# Population data
StatePop <- merge(df_pop_state, StateCenter, by="region") 
head(StatePop, 1)
```

## Data Preprocessing
```{r warning=F, message=F}
# Value to point
PopPoint <- NULL 
for(i in 1:nrow(StatePop)){
    # 1M people -> one point
    for(j in 1:(StatePop[i, "value"] / 1000000)){
        PopPoint <- rbind(PopPoint, StatePop[i, ])   
    }
}
head(PopPoint, 3)
```

## Plotting
```{r warning=F, message=F}
USMap <- get_map(location = "United States", zoom = 4)
densityMap <- ggmap(USMap, extent = "device") + 
  geom_density2d(data = PopPoint, aes(x = lon, y = lat), size = 0.3) +
  stat_density2d(data = PopPoint, 
                 aes(x = lon, y = lat, fill = ..level.., alpha = ..level..), 
                 size = 0.01, bins = 16, geom = "polygon") + 
  scale_fill_gradient(low = "green", high = "red", guide = FALSE) + 
  scale_alpha(range = c(0, 0.3), guide = FALSE)
```

## Density Map
```{r warning=F, fig.height=5.5}
densityMap
```

## `ggmap` Reference
- [ggmap package source code](https://github.com/dkahle/ggmap)
- [ggmap cheat sheet](https://www.nceas.ucsb.edu/~frazier/RSpatialGuides/ggmap/ggmapCheatsheet.pdf)
- [ggmap doc](https://dl.dropboxusercontent.com/u/24648660/ggmap%20useR%202012.pdf)

# Take Home Message

## Grammar
- Data `ggplot(Data)`
- Geometries `geom_line()`, `geom_bar()`, `geom_point()`
- Aesthetic properties `ggplot(Data, aes(x=1, y=1, color='black'))`
- Scale mappings `scale_fill_continuous()`, `scale_fill_grey()`
- Stats `stat_density2d()`

## Reference
- [ggplot2 Cheat Sheet](https://www.rstudio.com/wp-content/uploads/2015/03/ggplot2-cheatsheet.pdf)
- Courtesy by Professor Yi-Ju Tseng, Chang Gung University, Taiwan